---
title: 'Connections Between Biodiversity and Redlined Neighborhoods'
author: "Carly Caswell"
output: html_document
date: "2023-10-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = TRUE, message = FALSE, warning = FALSE)
```

## Background

Present-day environmental justice may reflect legacies of injustice in the past. The United States has a long history of racial segregation which is still visible. During the 1930's the Home Owners' Loan Corporation (HOLC), as part of the New Deal, rated neighborhoods based on their perceived safety for real estate investment. Their ranking system, (A (green), B (blue), C (yellow), D (red)) was then used to block access to loans for home ownership. Colloquially known as "redlining", this practice has had widely-documented consequences not only for community wealth, but also health.^[Gee, G. C. (2008). A multilevel analysis of the relationship between institutional and individual racial discrimination and health status. American journal of public health, 98(Supplement_1), S48-S56.] Redlined neighborhoods have less greenery^[Nardone, A., Rudolph, K. E., Morello-Frosch, R., & Casey, J. A. (2021). Redlines and greenspace: the relationship between historical redlining and 2010 greenspace across the United States. Environmental health perspectives, 129(1), 017006.] and are hotter than other neighborhoods.^[Hoffman, J. S., Shandas, V., & Pendleton, N. (2020). The effects of historical housing policies on resident exposure to intra-urban heat: a study of 108 US urban areas. Climate, 8(1), 12.] 

Check out coverage by the [New York Times](https://www.nytimes.com/interactive/2020/08/24/climate/racism-redlining-cities-global-warming.html).

A recent study found that redlining has not only affected the environments communities are exposed to, it has also shaped our observations of biodiversity.^[Ellis-Soto, D., Chapman, M., & Locke, D. H. (2023). Historical redlining is associated with increasing geographical disparities in bird biodiversity sampling in the United States. Nature Human Behaviour, 1-9.] Community or citizen science, whereby individuals share observations of species, is generating an enormous volume of data. Ellis-Soto and co-authors found that redlined neighborhoods remain the most undersampled areas across 195 US cities. This gap is highly concerning, because conservation decisions are made based on these data.

Check out coverage by [EOS](https://eos.org/articles/bird-biodiversity-reports-reflect-cities-redlined-past?utm_source=EosBuzz&mkt_tok=OTg3LUlHVC01NzIAAAGOx3ckQPpeDY3QyFKdUnBSYqILj3qLrgb_Q2lasphwcF90xEzMgjZFUo3NHsw3caXDcZe6I8D8qdVsSRr3wzhb6LKAQRgNo8ffhjieEbDwMw).

## Goals

1.
2.
3.
4.

## The Data

#### Dataset 1: EJScreen
Data from the United States Environmental Protection Agency's EJScreen: Environmental Justice Screening and Mapping Tool.
According to the [US EPA website](https://www.epa.gov/ejscreen/purposes-and-uses-ejscreen):

>This screening tool and data may be of interest to community residents or other stakeholders as they search for environmental or demographic information. It can also support a wide range of research and policy goals. The public has used EJScreen in many different locations and in many different ways.
>
>EPA is sharing EJScreen with the public:  
> - to be more transparent about how we consider environmental justice in our work,  
> - to assist our stakeholders in making informed decisions about pursuing environmental justice and,   
> - to create a common starting point between the agency and the public when looking at issues related to environmental justice. 

EJScreen provides on environmental and demographic information for the US at the Census [tract](https://en.wikipedia.org/wiki/Census_tract) and [block group](https://en.wikipedia.org/wiki/Census_block_group) levels. I worked with block group data that has been downloaded from the [EPA site](https://www.epa.gov/ejscreen/download-ejscreen-data). Additional information, including limitations and caveats, can be explored here: [limitations and caveats](https://www.epa.gov/ejscreen/limitations-and-caveats-using-ejscreen).

#### Dataset 2: Mapping Inequality
A team of researchers, led by the [Digital Scholarship Lab](https://dsl.richmond.edu/) at the University of Richmond have digitized maps and information from the HOLC as part of the [Mapping Inequality](https://dsl.richmond.edu/panorama/redlining/#loc=5/39.1/-94.58) project. 

I worked with maps of HOLC grade designations for Los Angeles. Information on the data can be found [here](https://dsl.richmond.edu/panorama/redlining/#loc=5/39.1/-94.58&text=downloads).^[Robert K. Nelson, LaDale Winling, Richard Marciano, Nathan Connolly, et al., “Mapping Inequality,” American Panorama, ed. Robert K. Nelson and Edward L. Ayers, accessed October 17, 2023, https://dsl.richmond.edu/panorama/redlining/] 

#### Dataset 3: Biodiversity observations
The [Global Biodiversity Information Facility](gbif.org) is the largest aggregator of biodiversity observations in the world. Observations typically include a location and date that a species was observed. I am working with observations of birds from 2021 onward.


## Workflow

#### Importing
First I needed to load in relevant packages.Then I'm going to import EJScreen data and filtering to Los Angeles County for this analysis. 

```{r }
library(sf)
library(spData)
library(tmap)
library(tidyverse)
library(rmapshaper)
library(smoothr)
```

```{r }
#Loading in ejscreen data:
ejscreen <- st_read(here::here("data", "EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb")) 

#Filtering to LA County:
la_county <- ejscreen %>%
 filter(CNTY_NAME %in% c("Los Angeles County"))

```

#### Analyzing Census Block Groups

I'm going to look at some preliminary variables of the EJScreen data. First, I'm specifically going to look at the Wastewater discharge index and see if I can indicate which census block groups are above the 95th percentile of national values. To do this, I will map LA county and included the wastewater discharge variable by adding a centroid.This will allow me to understand my data a bit better and determine where the wastewater discharge is most concentrated.

```{r }
#First creating a centroid of the LA County data and filtering for census block groups that are greater than the 95th wastewater discharge percentile
la_percentile_95 <- st_centroid(la_county) %>% 
  filter(P_PWDIS > 95)


#Creating a map of the LA county data with the centroid for wastewater:
ggplot() +
  geom_sf(data = la_county, aes(fill = PWDIS)) +
  geom_sf(data = la_percentile_95, color = "red") +
  scale_fill_viridis_c() +  
  theme_minimal() 

```
Find the percent of census block groups that have:  
- less than 5% of the population is considered low income
(5 points)

```{r }
#Notes for me:
#P_LOWINCPCT = Percentile for % low income
la_county_lowinc <- la_county %>%
  filter(LOWINCPCT < .05) %>% 
  nrow()/nrow(LA_ejscreen) *100

```

Find the percent of census block groups that are:    
- above the 80th percentile for Particulate Matter 2.5 AND  
- above the 80th percentile for Superfund proximity
(10 points)

```{r }
#Notes for me:
#P_PM25 =	Percentile for Particulate Matter 2.5
#P_PNPL	Percentile for Superfund proximity

la_county_pm <- la_county %>%
  filter(P_PM25 > 80)

cbg_ex1 <- la_county_pm %>% 
  filter(P_PNPL > 80)


 (nrow(cbg_ex1) / nrow(la_county)) * 100

#17.357%
```

Import redlining information for Los Angeles.
```{r }
LA_redlining <- st_read("https://dsl.richmond.edu/panorama/redlining/static/downloads/geojson/CALosAngeles1939.geojson") %>%
  st_make_valid()

```

Make a map of historical redlining boundaries, colored by HOLC grade.
(5 points)
```{r }
ggplot() +
  geom_sf(data = la_county) +
  geom_sf(data = LA_redlining, aes(color = holc_grade)) +
  scale_fill_viridis_c() +  
  theme_minimal()

```

Find the number of census block groups that fall within areas with HOLC grades
hint: make sure the CRS match
(15 points)
```{r eval=FALSE}
la_crs <- st_crs(la_county)
la_redlining_crs <- st_crs(LA_redlining)

# I found that the two CRS' are different, so I changed them to match:
la_redlining2 <- st_transform(LA_redlining, crs = la_crs)

#confirmed with the below statement that the CRS' are the same:
la_crs == st_crs(la_redlining2)

#Combined data
holc_grade_join <- st_join(la_county, la_redlining2, join = st_within)

# Filtered for the census block groups that fell within areas with HOLC grades:
census_block_groups_in_holc <- holc_grade_join %>% 
  filter(!is.na(holc_grade))

# Counted the number of census block groups:
nrow(census_block_groups_in_holc)

#There are 887 census block groups
```

Summarize current conditions based on EJScreen data within historical redlining categories using the mean of the following variables:  
-% low income.  
- percentile for particulate Matter 2.5.  
- percentile for low life expectancy.  
- percentile for air toxics cancer risk
(20 points)

```{r }
summary_redlining_data <- holc_grade_join %>%
  group_by(holc_grade) %>%
  summarize(mean_perclow_income = mean(LOWINCPCT, na.rm = TRUE),
            mean_pm_2.5_percentile = mean(P_PM25, na.rm = TRUE),
            mean_low_life_expectancy_percentile = mean(P_LIFEEXPPCT, na.rm = TRUE),
            mean_air_toxics_cancer_risk_percentile = mean(P_CANCER, na.rm = TRUE))

# Printing the summarized data
print(summary_redlining_data)



```

Please write 2-3 sentences reflecting on these results.
(5 points)

Grade C seems to have the highest means of the groups of percentiles, indicating this holc grade of LA county has higher % of low income, pm 2.5, low life expectancy, and higher air toxic cancer risks.

Grade A seems to consistently have the lowest means of the groups of percentiles, indicating this holc grade has a lower risk of air toxins, cancer risk, pm 2.5, and fewer are classified as lower income. 

### Investigate the legacy of redlining in biodiversity observations

For bird observations from 2022 that fall within neighborhoods with HOLC grades, find the percent of observations within each redlining categories and plot results.
hint: make sure that the bird observations have the same CRS as redlining data.
(20 points)

**-5 wrong order of intersect**

````{r }
birds <- st_read(here::here("data", "gbif-birds-LA"))

#Does the birds crs match the la redlining crs?
st_crs(birds) == st_crs(la_redlining2)

# I found that the two CRS' are different, so I changed them to match:
birds2 <- st_transform(birds, crs = st_crs(la_redlining2))

#Confirmed with the below statement that the CRS' are the same:
st_crs(birds2) == st_crs(la_redlining2)

#Combined data
holc_grade_bird_join <- st_join(la_redlining2, birds2, join = st_within)

# Calculate the percentage of bird observations within each redlining category
percentage_data <- holc_grade_bird_join %>%
  group_by(holc_grade) %>%
  summarise(Percentage = (n() / nrow(birds2)) * 100)

# Plot the results
ggplot(percentage_data, aes(x = holc_grade, y = Percentage, fill = holc_grade)) +
  geom_bar(stat = "identity") +
  labs(x = "HOLC Grade", y = "Percentage of Bird Observations", title = "Percentage of Bird Observations within Redlining Categories") +
  theme_minimal()
```

## Conclusion

Grade C, the grade with the higher % of low income, pm 2.5, low life expectancy, and higher air toxic cancer risks has the highest percentage of birds in the area.Grade A, the one with the lower risk of air toxins, cancer risk, pm 2.5, and fewer are classified as lower income has the lowest percentage of birds in the area.This is surprising because I would expect neighborhoods with higher incomes to potentially have more investment in parks, lawns, recreation areas that would facilitate biodiversity.One of the implications for the low income communities could be less infrastructure for waste control, which could be a factor in the larger percentage of birds in those areas.





