---
title: 'Connections Between Biodiversity and Redlined Neighborhoods'
author: "Carly Caswell"
output: html_document
date: "2023-10-17"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = TRUE, message = FALSE, warning = FALSE)
```

## Background

Present-day environmental justice may reflect legacies of injustice in
the past. The United States has a long history of racial segregation
which is still visible. During the 1930's the Home Owners' Loan
Corporation (HOLC), as part of the New Deal, rated neighborhoods based
on their perceived safety for real estate investment. Their ranking
system, (A (green), B (blue), C (yellow), D (red)) was then used to
block access to loans for home ownership. Colloquially known as
"redlining", this practice has had widely-documented consequences not
only for community wealth, but also health.[^1] Redlined neighborhoods
have less greenery[^2] and are hotter than other neighborhoods.[^3]

[^1]: Gee, G. C. (2008). A multilevel analysis of the relationship
    between institutional and individual racial discrimination and
    health status. American journal of public health, 98(Supplement_1),
    S48-S56.

[^2]: Nardone, A., Rudolph, K. E., Morello-Frosch, R., & Casey, J. A.
    (2021). Redlines and greenspace: the relationship between historical
    redlining and 2010 greenspace across the United States.
    Environmental health perspectives, 129(1), 017006.

[^3]: Hoffman, J. S., Shandas, V., & Pendleton, N. (2020). The effects
    of historical housing policies on resident exposure to intra-urban
    heat: a study of 108 US urban areas. Climate, 8(1), 12.

Check out coverage by the [New York
Times](https://www.nytimes.com/interactive/2020/08/24/climate/racism-redlining-cities-global-warming.html).

A recent study found that redlining has not only affected the
environments communities are exposed to, it has also shaped our
observations of biodiversity.[^4] Community or citizen science, whereby
individuals share observations of species, is generating an enormous
volume of data. Ellis-Soto and co-authors found that redlined
neighborhoods remain the most undersampled areas across 195 US cities.
This gap is highly concerning, because conservation decisions are made
based on these data.

[^4]: Ellis-Soto, D., Chapman, M., & Locke, D. H. (2023). Historical
    redlining is associated with increasing geographical disparities in
    bird biodiversity sampling in the United States. Nature Human
    Behaviour, 1-9.

Check out coverage by
[EOS](https://eos.org/articles/bird-biodiversity-reports-reflect-cities-redlined-past?utm_source=EosBuzz&mkt_tok=OTg3LUlHVC01NzIAAAGOx3ckQPpeDY3QyFKdUnBSYqILj3qLrgb_Q2lasphwcF90xEzMgjZFUo3NHsw3caXDcZe6I8D8qdVsSRr3wzhb6LKAQRgNo8ffhjieEbDwMw).

## Goals

1.  Understand population characteristics and environmental factors of
    LA County's Census Block Groups

2.  Analyze population characteristics and environmental factors of
    redlined communities in LA County

3.  Assess 2022 bird populations in different HOLC grades of LA County

## The Data

#### Dataset 1: EJScreen

Data from the United States Environmental Protection Agency's EJScreen:
Environmental Justice Screening and Mapping Tool. According to the [US
EPA website](https://www.epa.gov/ejscreen/purposes-and-uses-ejscreen):

> This screening tool and data may be of interest to community residents
> or other stakeholders as they search for environmental or demographic
> information. It can also support a wide range of research and policy
> goals. The public has used EJScreen in many different locations and in
> many different ways.
>
> EPA is sharing EJScreen with the public:\
> - to be more transparent about how we consider environmental justice
> in our work,\
> - to assist our stakeholders in making informed decisions about
> pursuing environmental justice and,\
> - to create a common starting point between the agency and the public
> when looking at issues related to environmental justice.

EJScreen provides on environmental and demographic information for the
US at the Census [tract](https://en.wikipedia.org/wiki/Census_tract) and
[block group](https://en.wikipedia.org/wiki/Census_block_group) levels.
I worked with block group data that has been downloaded from the [EPA
site](https://www.epa.gov/ejscreen/download-ejscreen-data). Additional
information, including limitations and caveats, can be explored here:
[limitations and
caveats](https://www.epa.gov/ejscreen/limitations-and-caveats-using-ejscreen).

#### Dataset 2: Mapping Inequality

A team of researchers, led by the [Digital Scholarship
Lab](https://dsl.richmond.edu/) at the University of Richmond have
digitized maps and information from the HOLC as part of the [Mapping
Inequality](https://dsl.richmond.edu/panorama/redlining/#loc=5/39.1/-94.58)
project.

I worked with maps of HOLC grade designations for Los Angeles.
Information on the data can be found
[here](https://dsl.richmond.edu/panorama/redlining/#loc=5/39.1/-94.58&text=downloads).[^5]

[^5]: Robert K. Nelson, LaDale Winling, Richard Marciano, Nathan
    Connolly, et al., "Mapping Inequality," American Panorama, ed.
    Robert K. Nelson and Edward L. Ayers, accessed October 17, 2023,
    <https://dsl.richmond.edu/panorama/redlining/>

#### Dataset 3: Biodiversity observations

The [Global Biodiversity Information Facility](gbif.org) is the largest
aggregator of biodiversity observations in the world. Observations
typically include a location and date that a species was observed. I am
working with observations of birds from 2021 onward.

## Workflow

### Preliminary Analysis

#### Importing

First I needed to load in relevant packages.Then I'm going to import
EJScreen data and filtering to Los Angeles County for this analysis.

```{r }
library(sf)
library(spData)
library(tmap)
library(tidyverse)
library(rmapshaper)
library(smoothr)
```

```{r }
#Loading in ejscreen data:
ejscreen <- st_read(here::here("data", "EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb")) 

#Filtering to LA County:
la_county <- ejscreen %>%
 filter(CNTY_NAME %in% c("Los Angeles County"))

```

#### Analyzing Census Block Groups

I'm going to look at some preliminary variables of the EJScreen data.
First, I'm specifically going to look at the Wastewater discharge index
and see if I can indicate which census block groups are above the 95th
percentile of national values. To do this, I will map LA county and
included the wastewater discharge variable by adding a centroid.This
will allow me to understand my data a bit better and determine where the
wastewater discharge is most concentrated.

```{r }
#First creating a centroid of the LA County data and filtering for census block groups that are greater than the 95th wastewater discharge percentile
la_percentile_95 <- st_centroid(la_county) %>% 
  filter(P_PWDIS > 95)


#Creating a map of the LA county data with the centroid for wastewater:
ggplot() +
  geom_sf(data = la_county, aes(fill = PWDIS)) +
  geom_sf(data = la_percentile_95, color = "red") +
  scale_fill_viridis_c() +  
  theme_minimal() 

```

Second, I'm going to look at the population considered low income in LA
county. To do this, I will use my LA county data and specifically filter
it to see those census block groups that include a low income population
that makes up less than 5% of each census group's population.This will
allow me to understand my data a bit better and determine where the
non-low income neighborhoods are most concentrated.

```{r }
#Filtering my data to only include neighborhoods with a small percentage of low income population:
la_county_lowinc <- la_county %>%
  filter(LOWINCPCT < .05) %>% 
  nrow()/nrow(LA_ejscreen) *100

print(paste('The percentage of LA county that includes less than 5% of low income population is:', la_county_lowinc))
```

Third, I'm going to look at the populations in LA county that are above
certain percentiles for environmental threats. One threat I'm going to
look at is particulate matter 2.5 (PM 2.5), which is a considerable
measure of air pollution that can cause adverse health effects.To do
this, I will use my LA county data and filter to areas that are above
the 80th percentile for PM 2.5. Another threat I'm going to look into is
superfund proximity. I will use the LA County data to filter to areas
that are above the 80th percentile for being near a superfund site,
which refers to the proximity or closeness of a location to an area that
is contanimated or hazardous.This will allow me to understand some
statistics on the population of LA County as it relates to environmental
concerns.

```{r }
#Filtering data:
la_county_pm <- la_county %>%
  filter(P_PM25 > 80)

cbg_ex1 <- la_county_pm %>% 
  filter(P_PNPL > 80)

print((nrow(cbg_ex1) / nrow(la_county)) * 100)

#17.357% of LA county is in the 80th percentile or greater for being close to superfund sites and exposure to Particulate Matter 2.5. 
```

### Redlining Analysis

#### Importing

Importing redlining information for Los Angeles County.

```{r }
LA_redlining <- st_read("https://dsl.richmond.edu/panorama/redlining/static/downloads/geojson/CALosAngeles1939.geojson") %>%
  st_make_valid()
```

#### Visualizing this Data

I first want to map the data I'm working with to understand the
redlining boundaries. I'm going to use the HOLC grades to see the
classified neighborhoods (Grades A through D) based on specific
socieoconomic categories.

```{r }
#Mapping LA county and the redlining data:
ggplot() +
  geom_sf(data = la_county) +
  geom_sf(data = LA_redlining, aes(color = holc_grade)) +
  scale_fill_viridis_c() +  
  theme_minimal()

```

#### Analyzing Redlined Neighborhoods

Now I want to combine my LA county census block groups with my HOLC
grades to see which neighborhoods were subject to redlining. To do this,
I'm going to first validate the CRS' of my redlining data and LA county
data match, and then I'm going to join the data to find where the
redlining grades lie within the larger LA County region. This will set
me up to compare redlined neighborhoods with my previous analysis of
neighborhood characteristics, such as being low income.

```{r eval=FALSE}
#Checking the crs' before I can join:
la_crs <- st_crs(la_county)
la_redlining_crs <- st_crs(LA_redlining)
## I found that the two CRS' are different, so I changed them to match:
la_redlining2 <- st_transform(LA_redlining, crs = la_crs)
##Confirming with the below statement that the CRS' are the same:
la_crs == st_crs(la_redlining2)


#Combined my redlining data and la county data 
holc_grade_join <- st_join(la_county, la_redlining2, join = st_within)

#Filtered for the census block groups that fell within areas with HOLC grades:
census_block_groups_in_holc <- holc_grade_join %>% 
  filter(!is.na(holc_grade))

#Counted the number of census block groups in my filtered area:
nrow(census_block_groups_in_holc)

#There are many redlined neighborhoods in LA county!
```

Now that I have HOLC grades included in my LA county region, I can
summarize characteristics of variables I was initially exploring in my
analysis. I'm going to calculate, for each grade, the means of the
following variables: % of low income population, percentile for PM 2.5,
percentile for low life expectancy, and percentile for air toxic cancer
risk in each grade

```{r }
summary_redlining_data <- holc_grade_join %>%
  group_by(holc_grade) %>%
  summarize(mean_perclow_income = mean(LOWINCPCT, na.rm = TRUE),
            mean_pm_2.5_percentile = mean(P_PM25, na.rm = TRUE),
            mean_low_life_expectancy_percentile = mean(P_LIFEEXPPCT, na.rm = TRUE),
            mean_air_toxics_cancer_risk_percentile = mean(P_CANCER, na.rm = TRUE))

# Printing the summarized data
print(summary_redlining_data)

```

I can see that Grade C seems to have the highest means of the groups of
percentiles, indicating this HOLC grade of LA county has a higher % of
low income population, pm 2.5, low life expectancy, and higher air toxic
cancer risks.Grade A seems to consistently have the lowest means of the
groups of percentiles, indicating this holc grade has a lower risk of
air toxins, cancer risk, pm 2.5, and fewer members of the population are
classified as lower income.

#### Analyzing redlining communities as they relate to biodiversity observations

Taking this a step further, I'm going to consider biodiversity and look
at bird observations in this area from 2022. I want to compare bird
populations in neighborhoods with HOLC grades to see if specific grades
have a trend in bird populations. To do this, I'm going to load in my LA
county bird data and find the percent of observations within each
redlining category.

```{r}
#Reading in my LA bird data 
birds <- st_read(here::here("data", "gbif-birds-LA"))

#Does the birds crs match the la redlining crs? st_crs(birds) ==
st_crs(la_redlining2)

#I found that the two CRS' are different, so I changed them to match:
birds2 <- st_transform(birds, crs = st_crs(la_redlining2))
#Confirming with the below statement that the CRS are the same:
st_crs(birds2) == st_crs(la_redlining2)

#Combining my data
holc_grade_bird_join <-st_join(birds2,la_redlining2, join = st_within)

#Calculating the percentage of bird observations within each redlining category

percentage_data <- holc_grade_bird_join %\>% group_by(holc_grade) %\>%
summarise(Percentage = (n() / nrow(birds2)) * 100)

#Plotting my results

ggplot(percentage_data, aes(x = holc_grade, y = Percentage, fill =
holc_grade)) + geom_bar(stat = "identity") + labs(x = "HOLC Grade", y =
"Percentage of Bird Observations", title = "Percentage of Bird
Observations within Redlining Categories") + theme_minimal()
```

## Conclusion

Grade C, the grade with the higher percentage of low income, Particulate
Matter 2.5, low life expectancy, and higher air toxic cancer risks has
the highest percentage of birds in the LA County area. Grade A, with the
lower risk of air toxins, cancer risk, Particulate Matter 2.5, and fewer
population that is classified as lower income has the lowest percentage
of birds in the area. This is surprising because I would expect
neighborhoods with higher incomes to potentially have more investment in
parks, lawns, or recreation areas that would facilitate biodiversity.
Knowing that Grade C has a higher risk of toxic cancer risks, air
pollution, and superfund proximity, this also concerns me regarding the
state of the bird species in LA County outside of the human impact. This
could be contributing to longer-term effects of decreasing bird
populations if these environmental conditions are prevalent in areas
where the bird populations are mostly residing.
